A new tidy data structure to support exploration and modeling of temporal data

### 2019-08-29 tsibble.Rmd
case studies and captions

A closer examination of some big airports across the US will give an indication of how well the busiest airports manage the outflow traffic on a daily basis. A subset that contains observations for Houston (IAH), New York (JFK), Kalaoa (KOA), Los Angeles (LAX) and Seattle (SEA) airports is obtained first. The succeeding operations compute delayed percentages every day at each airport, which are framed as grey lines in Figure \@ref(fig:sel-monthly-plot). Winter months tend to fluctuate a lot compared to the summer across all the airports. Superimposed on the plot are two-month moving averages, so the temporal trend is more visible. Since the number of days for each month is variable, moving averages over two months will require weights input. But the weights specification can be avoid using a pair of commonly used rectangling verbs--`nest()` and `unnest()`, to wrap data frames partitioned by months into list-columns. The sliding operation with a large window size smooths out the fluctuations and gives a stable trend around 25% over the year. LAX airport has seen a gradual decline in delays over the year, whereas the SEA airport has a steady number delays over time. The IAH and JFK airports have more delays in the middle of year, while the KOA has the inverse pattern with higher delay percentage in both ends of the year. This pipeline gets the data into the daily series, and shifts the focus to five major airports.

@@

(ref:sel-monthly-plot-cap) Daily delayed percentages for departure with two-month moving averages overlaid at five international airports. There are least fluctuations and relatively fewer delays observed at KOA airport. The estimates of temporal trend are around 25% across the other four airports, but highlight different time periods of severe delays.

@@

(ref:count-gaps-cap) Time gap plots for the 49 customers with most implicit missing values, and the remaining customers grouped into the one line at the bottom panel. Each cross represents an observation missing in time and a line between two dots shows continuous missingness over time. Each row corresponds to one customer. Missing values occur at various times, with more in January and February than other months.

Aggregation across all individuals helps to sketch a big picture of the behavioral change over time in the region, organized into a calendar display (Figure \@ref(fig:calendar-plot)) using the **sugrrants** package [@wang-calendarviz-2018]. Each glyph represents the daily pattern of average residential electricity usage every thirty minutes. Higher consumption is indicated by higher values, and typically occurs in daylight hours. Color indicates hot days. The daily snapshots vary depending on the season in the year. During the summer months (December and January), the late-afternoon peak becomes predominant driven by the use of air conditioning, especially on hot days with daily average temperature greater than 25 degrees Celsius. However, the winter time (July and August) sees two peaks in a day, which is probably due to heating in the morning and evening.

@@

(ref:calendar-plot-cap) Half-hourly average electricity use across all customers in the region, organized into calendar format, with color indicating hot days. Energy use of hot days tends to be higher, suggesting air conditioner use. Days in the winter months have a double peak suggesting heater use.

Lastly, a common practice in the energy sector is load forecasting. Forecasting the demand for the last day of 2013 is undertaken, holding the original energy data as the validation set. The most recent December data is fitted by ARIMA models with and without temperature information using automatic order selection [@autoarima]. The logarithmic transformation is applied to the average demand to ensure the positive forecasts. Figure \@ref(fig:elec-forecast) plots the one-day forecasts obtained from both models against the actual demand within the last two-week window. ARIMA regressed by average temperatures gives a better fit than the one without, although both tend to underestimate the night demand. The forecasting performance is reported in Table \@ref(tab:elec-accuracy), consistent with the findings in Figure \@ref(fig:elec-forecast).

@@

(ref:elec-forecast-cap) One-day forecasts generated by ARIMA models with and without temperatures plotted against the actual demand. Both nicely capture the temporal dynamics, but ARIMA with temperatures performs better than the one without.

>>>>

A closer examination of some big airports across the US will give an indication of how well the busiest airports manage the outflow traffic on a daily basis. A subset that contains observations for Houston (IAH), New York (JFK), Kalaoa (KOA), Los Angeles (LAX) and Seattle (SEA) airports is obtained first. The succeeding operations compute delayed percentages every day at each airport, which are shown as grey lines in Figure \@ref(fig:sel-monthly-plot). Winter months tend to fluctuate a lot compared to the summer across all the airports. Superimposed on the plot are two-month moving averages, so the temporal trend is more visible. Since the number of days for each month is variable, moving averages over two months will require a weights input. But the weights specification can be avoided using a pair of commonly used rectangling verbs--`nest()` and `unnest()`, to wrap data frames partitioned by months into list-columns. The sliding operation with a large window size smooths out the fluctuations and gives a stable trend around 25% over the year, for IAH, JFK, LAX and SEA. LAX airport has seen a gradual decline in delays over the year, whereas the SEA airport has a steady delay. The IAH and JFK airports have more delays in the middle of year, while the KOA has the inverse pattern with higher delay percentage in both ends of the year. This pipeline gets the data into the daily series, and shifts the focus to five major airports.

@@

(ref:sel-monthly-plot-cap) Daily delayed percentages for flight departure, with two-month moving averages overlaid, at five international airports. There are least fluctuations, and relatively fewer delays, observed at KOA airport.

@@

(ref:count-gaps-cap) Exploring temporal location of missing values, using time gap plots for the 49 customers with most implicit missing values.  The remaining customers are grouped into the one line in the bottom panel. Each cross represents an observation missing in time and a line between two dots shows continuous missingness over time.  Missing values tend to occur at various times, although there is a higher concentration of missing in January and February for most customers.

Aggregation across all individuals helps to sketch a big picture of the behavioral change over time in the region, organized into a calendar display (Figure \@ref(fig:calendar-plot)) using the **sugrrants** package [@wang-calendarviz-2018]. Each glyph represents the daily pattern of average residential electricity usage every thirty minutes. Higher consumption is indicated by higher values, and typically occurs in daylight hours. Color indicates hot days. The daily snapshots vary depending on the season in the year. During the summer months (December and January), the late-afternoon peak becomes the dominant usage pattern. This is probably driven by the use of air conditioning, because high peaks mostly correspond to hot days, where daily average temperature greater than 25 degrees Celsius. In the winter time (July and August) the daily pattern sees two peaks, which is probably due to heating in the morning and evening.

@@

(ref:calendar-plot-cap) Half-hourly average electricity use across all customers in the region, organized into calendar format, with color indicating hot days. Energy use of hot days tends to be higher, suggesting air conditioner use. Days in the winter months have a double peak suggesting morning and evening heater use.

A common practice with energy data analysis is load forecasting, because providers need to know they have capacity to supply electricity. To illustrate the pipeline including modeling, here  demand is modeled for December 2013, with the usage forecast for the last day (48 steps ahead because the data is half-hourly). The energy data for the last day is not used for modeling.   ARIMA models with and without temperature information are fitted using automatic order selection [@autoarima]. The logarithmic transformation is applied to the average demand to ensure the positive forecasts. Figure \@ref(fig:elec-forecast) plots the one-day forecasts from both models against the actual demand, for the last two-week window. ARIMA regressed by average temperatures gives a better fit than the one without, although both tend to underestimate the night demand. The forecasting performance is reported in Table \@ref(tab:elec-accuracy), consistent with the findings in Figure \@ref(fig:elec-forecast).

@@

(ref:elec-forecast-cap) One-day (48 step ahead) forecasts generated by ARIMA models, with and without temperatures, plotted against the actual demand. Both nicely capture the temporal dynamics, but ARIMA with temperature performs better than the model without.
